Technical Dimensions of Programming Systems

### 2022-05-27 prog-2022/prog22-master.tex
copyedit, tweak etc.

In computer science, much has been said and done to allows comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems}; we believe that programming systems deserve a theory too. We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them. We identify technical dimensions by examining past influential programming systems and reviewing their design principles, technical capabilities, and styles of user interaction.

Technical dimensions capture characteristics that may be studied, compared and advanced indepently. This makes it possible to talk about programming systems in a way that can be shared and constructively debated rather than relying solely on personal impressions. We outline two concrete ways of using our framework. First, we show how to use it to analyze a recently developed novel programming system, and, second, we use it to identify an interesting unexplored point in the design space of programming systems. By providing foundations for more systematic research, we can help programming systems researchers to stand, at last, on the shoulders of giants.

>>>>

In computer science, much has been said and done to allow comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems}; we believe that programming systems deserve a theory too. We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them. We identify technical dimensions by examining past influential programming systems and reviewing their design principles, technical capabilities, and styles of user interaction.

Technical dimensions capture characteristics that may be studied, compared and advanced independently. This makes it possible to talk about programming systems in a way that can be shared and constructively debated rather than relying solely on personal impressions. We outline two concrete ways of using our framework. First, we show how to use it to analyze a recently developed novel programming system, and, second, we use it to identify an interesting unexplored point in the design space of programming systems. By providing foundations for more systematic research, we can help programming systems researchers to stand, at last, on the shoulders of giants.
